---
title: "Combining Species Presence/Absence with Environmental Data"
author: "Eleanor (Ella) Crotty"
header-includes: \usepackage{fvextra} \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
  word_document:
    toc: yes
urlcolor: blue
---

# Setup 

```{r Package Imports, message = F, warning = F, results = F}
# Warnings and startup messages suppressed
library(tidyverse)
library(patchwork)
library(scales)
library(ggrepel)
library(ggbreak)
library(readxl)
library(here)
```

```{r, results = F, message = F}
# Import data
SpeciesDetections <- read_csv(here("OCNMS_Project", "Outputs", "SOI_IDs_Species10Detections.csv")) %>% 
  filter(year(Date_UTC) != 2023)
EnvData1 <- read_csv(here("OCNMS_Project", "Data", "EnvironmentalDataset1.csv")) %>%  # Using EnvironmentalDataset1 because the satellite + NEMO data in EnvironmentalDataset2 didn't turn out to be good at gap filling
  filter(year != 2023) %>%  # Ignoring 2023 due to gaps for now
  mutate(year = as.factor(year))
SamplingDates2 <- read_csv(here("OCNMS_Project", "Data", "SamplingDates2.csv")) %>%  # Exported from EnvironmentalDataxSampleDates.Rmd, simple dataframe of all datetimes of samples
  filter(year(Date) != 2023) %>% # Ignoring 2023 due to gaps for now
  mutate(Source = case_when(Source == "Bottle_DNA" ~ "Bottle_DNA_Sampled", Source == "PPS_DNA" ~ "Automated_DNA_Sampler"))
```
 # Date check
 
```{r Checking Dates, eval = F}
# Let's take a quick look at whether the dates match up
sampdates <- data.frame(Date = SamplingDates2$Date, Source = "Metadata")
detectdates <- data.frame(Date = SpeciesDetections$Date_UTC, Source = "Detections")
datescomp <- rbind(sampdates,detectdates)

ggplot() +
  geom_vline(data = datescomp, mapping = aes(xintercept = Date, color = Source, linetype = Source)) +
  scale_linetype_manual(values = c(2,1)) +
  theme_bw() +
  facet_wrap(facets = vars(year(Date)), scales = "free_x", ncol = 1) #+
  #scale_x_datetime(date_breaks = "4 days", date_labels = "%y-%b-%d") # This is fucked but tbh I don't need it
```

# Sampling x environmental data graph

```{r Recreate the Sampling x Environmental Data graph}
ggplot(EnvData1, aes(x = date, y = DO, color = source, alpha = source)) +
  geom_point() +
  scale_alpha_manual(values = c(1,0.1)) + # doesn't currently do anything
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = SamplingDates2, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("purple1", "black", "firebrick2", "dodgerblue2")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(title = "Dissolved Oxygen Data + Sampling Dates", caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen, vertical lines = DNA sampling times", alpha = "Environmental Data", color = "Source")
```

# Data Join

```{r Data Join, results = F, message = F}
EnvRd <- EnvData1 %>% 
  mutate(DateMatch = round_date(date, unit = "10 minutes")) # Round to the nearest 10 minutes? 
DetectRd <- SpeciesDetections %>% 
  mutate(DateMatch = round_date(Date_UTC, unit = "10 minutes"), Date_local_hr = round_date(Date_local, unit = "hour")) # Spot check - looks good. 

eDNAxEnvData <- left_join(DetectRd, EnvRd, by = join_by(DateMatch)) 

investigate <- eDNAxEnvData %>% select(Species, DateMatch, Date_UTC, Date_local_hr, source, temperature, DO, SampleId, Rosette_position, Amplicon)
```

# Joined data graphs

```{r Graph eDNAxEnvData, fig.show = 'hide'}
# Currently hidden because these aren't super helpful
ggplot(eDNAxEnvData, aes(x = DateMatch, y = DO, shape = Present, size = Present, color = Present)) +
  scale_shape_manual(values = c(1, 19)) +
  scale_size_manual(values = c(1,1)) +
  geom_point() +
  theme_bw() +
  facet_wrap(facets = vars(Species))

ggplot(data = EnvData1, aes(x = date, y = DO)) +
  geom_line(color = "gray90") +
  geom_point(data = eDNAxEnvData, aes(x = DateMatch, y = DO, shape = Present, size = Present, color = Present)) +
  scale_shape_manual(values = c(1, 19)) +
  scale_size_manual(values = c(1,1)) + 
  theme_bw() +
  facet_wrap(facets = vars(Species, year(date)), scales = "free_x", ncol = 2)
```

```{r Preliminary Graphs}
EnvDataRange <- EnvData1 %>% 
  filter(date > as.POSIXct("2021-08-15 00:00:00", tz = "UTC"))

ggplot(data = EnvDataRange, aes(x = date, y = DO)) +
  geom_line(color = "gray90") +
  geom_point(data = eDNAxEnvData, aes(x = DateMatch, y = DO, shape = Present, size = Present, color = Present)) +
  scale_shape_manual(values = c(1, 19)) +
  scale_size_manual(values = c(1,1)) + 
  theme_bw() +
  facet_wrap(facets = vars(Species, year(date)), scales = "free_x", ncol = 4) +
  scale_x_datetime(breaks = "month", date_labels = "%b-%y") +
  theme(text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  geom_hline(yintercept = 2, linetype = 2) +
  labs(title = "Species Presence and Absence Over Dissolved Oxygen", x = "Date", y = "Dissolved Oxygen (mg/L) At TH042")

ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_Oxygen_Preliminary.png"), width = 2500, height = 2000, units = "px")

ggplot(data = EnvDataRange, aes(x = date, y = temperature)) +
  geom_line(color = "gray90") +
  geom_point(data = eDNAxEnvData, aes(x = DateMatch, y = temperature, shape = Present, size = Present, color = Present)) +
  scale_shape_manual(values = c(1, 19)) +
  scale_size_manual(values = c(1,1)) + 
  theme_bw() +
  facet_wrap(facets = vars(Species, year(date)), scales = "free_x", ncol = 4) +
  scale_x_datetime(breaks = "month", date_labels = "%b-%y") +
  theme(text = element_text(size = 15), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Species Presence and Absence Over Temperature", x = "Date", y = "Temperature (C) At TH042")

ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_Temp_Preliminary.png"), width = 2500, height = 2000, units = "px")
```

+
  annotate(geom = "label",
    x = hypoxiaDatetimes,
    y = rep(3, 6), # 3 mg/L, 6 times
    label = hypoxiaDates,
    vjust = rep(2, 6))
    
# Hypoxic Events
```{r}
EnvDataRange <- EnvDataRange %>% 
  mutate(dayLocal = as.Date(date, tz = "America/Los_Angeles"))
  
Hypoxia <- EnvDataRange %>% 
  filter(DO < 2) %>% 
  relocate(dayLocal, .after = date)

dayLocalHypoxia <- EnvDataRange %>% 
  mutate(dayLocal = as.Date(date)) %>% 
  relocate(dayLocal, .after = date) %>% 
  group_by(source, dayLocal, year) %>% 
  summarize(avTemp = mean(temperature), avDO = mean(DO)) %>% 
  filter(avDO < 2) 

# For threshold 2 mg/L:
# Approx. 2021-08-15 to 08-31
# Approx. 2021-09-20 to 09-26
# Approx. 2022-08-11 to 09-20

# Manually enter the starts & ends of dates
hypoxiaDates <- c("2021-08-15", "2021-08-31", "2021-09-20", "2021-09-26", "2022-08-11", "2022-09-20")
hypoxiaDatetimes <- as.POSIXct(hypoxiaDates, tz = "America/Los_Angeles") # as.POSIXct for graphing
hypoxiaLines <- data.frame(Label = hypoxiaDates, Date = hypoxiaDatetimes) %>% 
  mutate(year = year(Date))

HypoxicEvents <- EnvDataRange %>% # Cut down the entire dataset to the hypoxic event dates from dayLocalHypoxia
  mutate(Event = case_when(dayLocal >= as.Date("2021-08-15") & dayLocal <= as.Date("2021-08-31") ~ 1,
                           dayLocal >= as.Date("2021-09-20") & dayLocal <= as.Date("2021-09-26") ~ 2,
                           dayLocal >= as.Date("2022-08-11") & dayLocal <= as.Date("2022-09-20") ~ 3))

HypoxicEventsSummary <- HypoxicEvents %>% # Take the average DO and temp, and also calculate lengths
  group_by(Event) %>% 
  summarize(Start = min(date), End = max(date), AvgOxygen = mean(DO, na.rm = T), AvgTemp = mean(temperature)) %>% 
  mutate(Length = End - Start)

ytext1 <- 0.5
ytext2 <- 0.2

ggplot(EnvDataRange, aes(x = date, y = DO)) +
  geom_line() + 
  theme_bw() +
  geom_hline(yintercept = 2, 
             color = "red", 
             linetype = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank()) +
  scale_x_datetime(breaks = "2 weeks", 
                   date_labels = "%y-%m-%d",
                   limits = c(as.POSIXct("2021-08-10"), as.POSIXct("2022-09-20"))) + # Need to extend a little for line label
  geom_point(data = Hypoxia, 
             aes(x = date, y = DO), 
             color = "red", 
             size = 0.1) +
  geom_vline(
    xintercept = hypoxiaDatetimes, 
    color = "red") +
  geom_text(data = hypoxiaLines, 
            aes(x=Date, y=7.5, label=Label),
            inherit.aes = T,
            vjust = -0.5,
            angle = 90) +
  annotate(geom = "text", x = as.POSIXct("2021-08-22 12:00:00"), y = ytext1, label = "16 days", size = 3) +
  annotate(geom = "text", x = as.POSIXct("2021-09-30 20:00:00"), y = ytext1, label = "<- 6 d", size = 3) +
  annotate(geom = "text", x = as.POSIXct("2022-08-30"), y = ytext1, label = "39 d", size = 3) +
  annotate(geom = "text", x = as.POSIXct("2021-08-22 20:00:00"), y = ytext2, label = "1.91 mg/L", size = 3) +
  annotate(geom = "text", x = as.POSIXct("2021-10-05 20:00:00"), y = ytext2, label = "1.75 mg/L", size = 3) +
  annotate(geom = "text", x = as.POSIXct("2022-08-30"), y = ytext2, label = "1.61 mg/L", size = 3) +
  scale_x_break(as.POSIXct(c("2021-10-12", "2022-05-24"))) +
  labs(title = "Lenghth & Severity of OCNMS Hypoxic Events", x = "Date", y = "Dissolved Oxygen (mg/L)", caption = "Hypoxic events were defined here as periods where the daily average DO was < 2 mg/L")

ggsave(here("OCNMS_Project", "Plots", "HypoxicEvents.png"))

ggplot(EnvDataRange, aes(x = date, y = DO)) +
  geom_line() + 
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x") +
  geom_hline(yintercept = 2, color = "red", linetype = 2) +
  scale_x_datetime(breaks = "2 weeks", date_labels = "%y-%m-%d") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_point(data = dayLocalHypoxia, aes(x = as.POSIXct(dayLocal), y = avDO), color = "red", size = 2)
```


# DO x Temp graphs

```{r, warning = F}
full <- ggplot(EnvDataRange, aes(x = temperature, y = DO, shape = source, color = as.factor(year))) +
  scale_shape_manual(values = c(4,19)) +
  scale_color_manual(values = c("blue", "black")) +
  geom_point(shape = 1, alpha = 0.5) +
  theme_bw() +
  labs(title = "DO vs Temp during sampling years")
ggsave(here("OCNMS_eDNA", "Plots", "DO_vs_Temp.png"), plot = full)

full_yr <- ggplot(EnvDataRange, aes(x = temperature, y = DO, shape = source, color = as.factor(year))) +
  scale_shape_manual(values = c(4,19)) +
  scale_color_manual(values = c("blue", "black")) +
  geom_point(shape = 1, alpha = 0.5) +
  theme_bw() +
  facet_wrap(facets = vars(year)) +
  labs(title = "DO vs Temp during sampling years")
ggsave(here("OCNMS_eDNA", "Plots", "DO_vs_TempYear.png"), plot = full_yr)

zoom <- ggplot(EnvDataRange, aes(x = temperature, y = DO, shape = source, color = as.factor(year))) +
  scale_shape_manual(values = c(4,19)) +
  scale_color_manual(values = c("blue", "black")) +
  geom_point(shape = 1, alpha = 0.5) +
  scale_x_continuous(limits = c(7, 10)) +
  scale_y_continuous(limits = c(0,6)) +
  theme_bw() +
  labs(title = "DO vs Temp during sampling years (Zoomed in)")
ggsave(here("OCNMS_eDNA", "Plots", "DO_vs_Temp_Zoomed.png"), plot = zoom)

zoom_yr <- ggplot(EnvDataRange, aes(x = temperature, y = DO, shape = source, color = as.factor(year))) +
  scale_shape_manual(values = c(4,19)) +
  scale_color_manual(values = c("blue", "black")) +
  geom_point(shape = 1, alpha = 0.5) +
  scale_x_continuous(limits = c(7, 10)) +
  scale_y_continuous(limits = c(0,6)) +
  theme_bw() +
  facet_wrap(facets = vars(year)) +
  labs(title = "DO vs Temp during sampling years (Zoomed in)")
ggsave(here("OCNMS_eDNA", "Plots", "DO_vs_Temp_ZoomedYear.png"), plot = zoom_yr)

(full + full_yr) / (zoom + zoom_yr)
```

# Binomial Regression

**Binomial Regression Assumptions**

1. Independent observations - *Check!*
2. No multicollinearity: Continuous independent variables should not be correlated with each other - *NOT MET - shouldn't do a DO x Temp x Presence model*
3. Binomial dependent variable, with two **mutually exclusive** values - *Check!*
4. Continuous independent variable - *Check!*
5. Linearity between the binomial and continuous variables - this is checked with a logit transformation - *See below, this varies by species*
6. Sufficient obeservations - the source I found recommends > 15 - *Check! 10 should be alright but I will note it in the results*
7. No outliers - *One outlier, which appears to be part of a real heatwave - I have not filtered it out yet*


## Check for outliers

```{r}
ggplot(EnvDataRange, aes(x = DO, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvData, mapping = aes(x = DO, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_TempxDO.png"), width = 2500, height = 2000, units = "px")
```

## Filter out outliers - currently not using

```{r}
eDNAxEnvData_forlm <- eDNAxEnvData %>% 
  filter(temperature < 11)
```

```{r}
ggplot(EnvDataRange, aes(x = DO, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvData_forlm, mapping = aes(x = DO, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_TempxDO.png"), width = 2500, height = 2000, units = "px")
```


## Check for linearity

"Assumption #5 involves the necessity of a linear relationship between the continuous independent variables and the logit transformation of the dependent variable. This linearity assumption implies that **for continuous independent variables** like income level, hours of exercise per week, and blood sugar levels, **there should be a linear relationship with the logit of the dependent variable**, such as the probability of developing diabetes. Various methods can be employed to assess this linearity, with one common approach being the **Box-Tidwell procedure.** This technique involves creating interaction terms between each continuous independent variable and its natural logarithm and adding these to the logistic regression model. This technique can be implemented using software like **SPSS Statistics, which offers the Binary Logistic procedure to test for this assumption.** The results of this test are then interpreted to decide the next steps in the analysis, depending on whether the linearity assumption holds or is violated. If the assumption is met, the analysis can proceed as planned. However, if the assumption is not met, adjustments to the model or alternative methods may be necessary to address the non-linearity appropriately." - [Binomial Logistic Regression](https://www.amstatisticalconsulting.com/binomial-logistic-regression/)


Basically, this equation needs to have a somewhat linear relationship with the independent variable: 

$$g(p) = log\left( \frac{p}{1-p} \right)$$
Where p = probability of 1 (if 1, p. if 0, 1-p)?

```{r}
joinSpeciesList <- split(eDNAxEnvData, eDNAxEnvData$Species) # Split by species

library(car) # Has a function for the Box-Tidwell procedure
?boxTidwell

joinSpeciesList <- lapply(joinSpeciesList, drop_na, DO)

# joinSpeciesList, glm, formula = Present ~ DO

# Single test
boxTidwell(Present ~ DO, other.x = ~ year, data = joinSpeciesList[[1]]) # other.x = any factors not to be transformed. i had to make year into a factor to make it accept this lmao, so I'm not sure this is valid!

# Loop to test all for linearity with temperature
for (i in 1:length(joinSpeciesList)) {
  print(paste(names(joinSpeciesList)[i], sep = " ", "Presence vs DO"))
  print(boxTidwell(Present ~ DO, other.x = ~ year, data = joinSpeciesList[[i]]))
}
# Loop to test all for linearity with DO
for (i in 1:length(joinSpeciesList)) {
  print(paste(names(joinSpeciesList)[i], sep = " ", "Presence vs Temp"))
  print(boxTidwell(Present ~ temperature, other.x = ~ year, data = joinSpeciesList[[i]]))
}
```

## Calculate regression

```{r}
oxmodels <- lapply(joinSpeciesList, glm, formula = Present ~ DO, family = "binomial") # Make the model for each species
lapply(oxmodels, summary) # Print the summary of each model

tmodels <- lapply(joinSpeciesList, glm, formula = Present ~ temperature, family = "binomial") # Make the model for each species (temperature)
lapply(tmodels, summary) # Print the summary of each model
```



S Xu#, M Chen#, T Feng, L Zhan, L Zhou, G Yu*. Use ggbreak to effectively utilize plotting space to deal with large datasets and outliers. Frontiers in Genetics. 2021, 12:774846. doi: 10.3389/fgene.2021.774846

