---
title: "eDNA Compared To Oxygen Saturation"
author: "Eleanor (Ella) Crotty"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    toc: TRUE
urlcolor: blue
---

```{r Package Imports, message = F, warning = F}
# Warnings and startup messages suppressed
library(tidyverse)
library(patchwork)
library(scales)
library(ggrepel)
library(ggpmisc)
library(readxl)
library(here)
library(gsw)
library(R.matlab)
library(ggbreak)
```

# Import data 

```{r}
allReads_Species <- read_csv(here("OCNMS_eDNA", "Outputs", "AllReads_Species.csv")) # Combined PCR replicates, combined ESVs in species
envData <- read_csv(here("OCNMS_Hypoxia", "Outputs", "EnvironmentalDataset1.csv")) %>% 
  filter(year != 2023) %>%  # Ignoring 2023 due to gaps for now
  mutate(year = as.factor(year))
```

```{r}
SpeciesDetect <- allReads_Species %>% 
  mutate(Present = case_when(TotalnReads == 0 ~ F, TotalnReads != 0 ~ T)) %>% 
  relocate(Present, .after = TotalnReads)
```

# Horrid MatLab Sidequest

Trying to figure out if the density units are correct

time: [4040x1 double] (MATLAB datenum)
dens: [4040x1 double] (sigma‐theta)

Is this the same as the envData potential_density?

```{r, eval = F}
TH21 <- readMat(here("OCNMS_Hypoxia", "Data", "2021.mat"))
str(TH21) # Data structure
```

This does say density is in sigma-theta

```{r, eval = F}
lst21 <- TH21$CT.TH042.210608.038
datenum21 <- c(lst21[[2]][[4]]) # datenums as vector/timeJ
dates21 <- as.POSIXct((datenum21 - 719529)*86400, origin = "1970-01-01", tz = "UTC") # subtract to get origin = January 0, 0000 (this is a made up date so as.POSIXct won't handle it natively), need to multiply by seconds in a day to go from datenum (days since made up Matlab epoch) to POSIXct (seconds since Unix epoch). Just kinda hoping they were using UTC here.
dens21 <- c(lst21[[2]][[6]]) # Density as vector
temp21 <- c(lst21[[2]][[1]]) # Temperature as vector

matCheck <- data.frame(dates21, dens21, temp21)
```

This did not match at all.



```{r Graph Checks}
# Just some sanity checks to make sure I have what I think I have
ggplot(data = envData, aes(x = date, y = DO, color = source)) +
  geom_point(size = 0.5) +
  facet_wrap(facets = vars(year), scales = "free_x") +
  theme_bw()

ggplot(data = SpeciesDetect, aes(x = Date_UTC)) +
  geom_histogram(bins = 100) +
  theme_bw()
```
# Calculate oxygen saturation state

[R package documentation for this](https://search.r-project.org/CRAN/refmans/gsw/html/gsw_O2sol_SP_pt.html)
Then in units of umol kg-1 compare [O2]measured / [O2]saturated *100% = saturation %.

**Usage**

gsw_O2sol_SP_pt(SP, pt)

**Arguments**
SP: Practical Salinity (PSS-78) [ unitless ]
pt: potential temperature (ITS-90) [ degC ]

**Value**
Oxygen solubility in micro-moles per kg. 

To go from O2 in ml/l to O2 in μmol/kg multiply by 44.660 and divide by density in CGS. [There is a small issue regarding what density to use, but it gets buried in the decimal place weeds. It is best to use sigma-0, i.e. (1000 + sigma0)/1000 in CGS.] - Converting O2 in volume units (ml/l) to O2 in mass units (μmol/kg) with Java OceanAtlas Custom Calculations (and also conversion of nutrients in μmol/l to μmol/kg), James h. Swift, UCSC Scripps Institution of Oceanography September 2021

1 mL/L = 1.42903 mg/L

## Evan says: 

Potential density is just sigma_theta+1000 (g kg-1; this is the definition of sigma compared to rho). That gives 1026.15 to 1026.18 in the table above (pretty close to the zero-order assumption!)

And in terms of the conversion, I always check the stoichiometric unit conversions: mg/L * mmol/mg * umol/mmol * L/g * g/kg -> mmol kg
So you don't need to convert to mL/L at all (units that only matter for titrations), just calculate the concentration directly from your input units.
Example: 3 mg/L * (mmol/31.9988 mg) * (1000 umol/mmol) * (L/1026g) * (1000 g/kg) = 3/31.9988/1026*1e6 = ~91 umol kg-1

$$
DO~\frac{mg}{L} \times \frac{mmol}{31.9988~mg} \times \frac{1000~\mu mol}{mmol} \times \frac{L}{1026 g} (density) \times \frac{1000 g}{kg}
$$

```{r}
envDataSat <- envData %>% 
  mutate(O2meas = DO/31.9988/(potential_density + 1000)*(1000^2), O2sat = gsw_O2sol_SP_pt(salinity, temperature), SatPct = (O2meas/O2sat)*100) %>% 
  relocate(c(O2meas, O2sat, SatPct), .after = DO)
```

# Join the data

```{r Data join}
EnvRd <- envDataSat %>% 
  mutate(DateMatch = round_date(date, unit = "10 minutes")) # Round to the nearest 10 minutes? 
DetectRd <- SpeciesDetect %>% 
  mutate(DateMatch = round_date(Date_UTC, unit = "10 minutes"), Date_local_hr = round_date(Date_local, unit = "hour")) # Spot check - looks good. 

eDNAxEnvDataSat <- left_join(DetectRd, EnvRd, by = join_by(DateMatch)) 

investigate <- eDNAxEnvDataSat %>% select(Species, DateMatch, Date_UTC, Date_local_hr, source, temperature, DO, E_no, Rosette_position, Amplicon)

system("say Data Join Complete")
```

# Graphs

## Presence

```{r}
# Filter for years of interest
eDNAxEnvDataSat <- eDNAxEnvDataSat %>% 
  filter(year %in% c(2021, 2022))
# Filter for species of interest
eDNAxEnvDataSat_SOI <- eDNAxEnvDataSat %>% 
  filter(Species %in% c("Acartia longiremis", "Calanus pacificus", "Clupea pallasii", "Oncorhynchus tshawytscha")) 

ggplot(data = envDataSat, aes(x = date, y = SatPct)) +
  geom_line(color = "gray90") +
  geom_point(data = eDNAxEnvDataSat_SOI, aes(x = DateMatch, y = SatPct, shape = Present, size = Present, color = Present)) +
  scale_shape_manual(values = c(1, 19)) +
  scale_size_manual(values = c(1,1)) + 
  theme_bw() +
  facet_wrap(facets = vars(Species, year(date)), scales = "free_x", ncol = 4) +
  scale_x_datetime(breaks = "month", date_labels = "%b-%y") +
  theme(text = element_text(size = 20), axis.text.x = element_text(angle = 45, hjust = 1), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  geom_hline(yintercept = 30, linetype = 2) +
  labs(title = "Species Presence and Absence Over Dissolved Oxygen", x = "Date", y = "Dissolved Oxygen (Saturation %) At TH042")

ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_Oxygen_pO2.png"), width = 2500, height = 2000, units = "px")
```
## Hypoxic Events

```{r}
EnvDataRange <- envDataSat %>% 
  filter(date > as.POSIXct("2021-08-15 00:00:00", tz = "UTC"))

EnvDataRange <- EnvDataRange %>% 
  mutate(dayLocal = as.Date(date, tz = "America/Los_Angeles"))
  
Hypoxia <- EnvDataRange %>% 
  filter(SatPct < 30) %>% 
  relocate(dayLocal, .after = date)

dayLocalHypoxia <- EnvDataRange %>% 
  mutate(dayLocal = as.Date(date)) %>% 
  relocate(dayLocal, .after = date) %>% 
  group_by(source, dayLocal, year) %>% 
  summarize(avTemp = mean(temperature), avDO = mean(SatPct)) %>% 
  filter(avDO < 30) 

# For threshold 30%:
# 2021-08-15 to 2021-09-26
# 2022-06-13 to 2022-06-16
# 2022-07-29 to 2022-09-20

start1 <- "2021-08-15"
end1 <- "2021-09-26"
start2 <- "2022-06-13"
end2 <- "2022-06-16"
start3 <- "2022-07-29"
end3 <- "2022-09-20"

# Manually enter the starts & ends of dates
hypoxiaDates <- c(start1, end1, start2, end2, start3, end3)
hypoxiaDatetimes <- as.POSIXct(hypoxiaDates, tz = "America/Los_Angeles") # as.POSIXct for graphing
hypoxiaLines <- data.frame(Label = hypoxiaDates, Date = hypoxiaDatetimes) %>% 
  mutate(year = year(Date))

HypoxicEvents <- EnvDataRange %>% # Cut down the entire dataset to the hypoxic event dates from dayLocalHypoxia
  mutate(Event = case_when(dayLocal >= as.Date(start1) & dayLocal <= as.Date(end1) ~ 1,
                           dayLocal >= as.Date(start2) & dayLocal <= as.Date(end2) ~ 2,
                           dayLocal >= as.Date(start3) & dayLocal <= as.Date(end3) ~ 3))

HypoxicEventsSummary <- HypoxicEvents %>% # Take the average DO and temp, and also calculate lengths
  group_by(Event) %>% 
  summarize(Start = min(date), End = max(date), AvgOxygen = mean(SatPct, na.rm = T), AvgTemp = mean(temperature)) %>% 
  mutate(Length = End - Start)

ytext1 <- 10
ytext2 <- 5
labelsize <- 3

ggplot(EnvDataRange, aes(x = date, y = SatPct)) +
  geom_line() + 
  theme_bw() +
  geom_hline(yintercept = 30, 
             color = "red", 
             linetype = 2) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank()) +
  scale_x_datetime(breaks = "2 weeks", 
                   date_labels = "%y-%m-%d",
                   limits = c(as.POSIXct("2021-08-10"), as.POSIXct("2022-09-20"))) + # Need to extend a little for line label
  geom_point(data = Hypoxia, 
             aes(x = date, y = SatPct), 
             color = "red", 
             size = 0.1) +
  geom_vline(
    xintercept = hypoxiaDatetimes, 
    color = "red") +
  geom_label(data = hypoxiaLines, 
            aes(x=Date, y=80, label=Label),
            inherit.aes = T,
            vjust = 0.5,
            angle = 90,
            size = 3) +
  annotate(geom = "text", x = as.POSIXct("2021-09-10 12:00:00"), y = ytext1, label = "43 days", size = labelsize) +
  annotate(geom = "text", x = as.POSIXct("2022-06-25 20:00:00"), y = ytext1, label = "<- 4 d", size = labelsize) +
  annotate(geom = "text", x = as.POSIXct("2022-09-01"), y = ytext1, label = "54 d", size = labelsize) +
  annotate(geom = "text", x = as.POSIXct("2021-09-10 20:00:00"), y = ytext2, label = "24%", size = labelsize) +
  annotate(geom = "text", x = as.POSIXct("2022-06-25 20:00:00"), y = ytext2, label = "27%", size = labelsize) +
  annotate(geom = "text", x = as.POSIXct("2022-09-01"), y = ytext2, label = "20%", size = labelsize) +
  scale_x_break(as.POSIXct(c("2021-10-12", "2022-05-24"))) +
  labs(title = "Lenghth & Severity of OCNMS Hypoxic Events", x = "Date", y = "Dissolved Oxygen (Saturation %)", caption = "Hypoxic events were defined here as periods where the daily average oxygen saturation was < 30%")

ggsave(here("OCNMS_Project", "Plots", "HypoxicEvents_pO2.png"))

system("say Hypoxic Event Graphs Saved")
```


# Binomial Regression

**Binomial Regression Assumptions**

1. Independent observations - *Check!*
2. No multicollinearity: Continuous independent variables should not be correlated with each other - *NOT MET - shouldn't do a DO x Temp x Presence model*
3. Binomial dependent variable, with two **mutually exclusive** values - *Check!*
4. Continuous independent variable - *Check!*
5. Linearity between the binomial and continuous variables - this is checked with a logit transformation - *See below, this varies by species*
6. Sufficient obeservations - the source I found recommends > 15 - *Check! 10 should be alright but I will note it in the results*
7. No outliers - *One outlier, which appears to be part of a real heatwave - I have not filtered it out yet*


## Check for outliers

```{r}
ggplot(EnvDataRange, aes(x = SatPct, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvDataSat_SOI, mapping = aes(x = SatPct, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
ggsave(filename = here("OCNMS_Project", "Plots", "SpeciesPresence_TempxDO.png"), width = 2500, height = 2000, units = "px")
```

## Filter out outliers - currently not using

```{r, eval = F}
eDNAxEnvDataSat_SOI <- eDNAxEnvDataSat_SOI %>% 
  filter(temperature < 11)
```

```{r}
ggplot(EnvDataRange, aes(x = SatPct, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvDataSat_SOI, mapping = aes(x = SatPct, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
```


## Check for linearity

"Assumption #5 involves the necessity of a linear relationship between the continuous independent variables and the logit transformation of the dependent variable. This linearity assumption implies that **for continuous independent variables** like income level, hours of exercise per week, and blood sugar levels, **there should be a linear relationship with the logit of the dependent variable**, such as the probability of developing diabetes. Various methods can be employed to assess this linearity, with one common approach being the **Box-Tidwell procedure.** This technique involves creating interaction terms between each continuous independent variable and its natural logarithm and adding these to the logistic regression model. This technique can be implemented using software like **SPSS Statistics, which offers the Binary Logistic procedure to test for this assumption.** The results of this test are then interpreted to decide the next steps in the analysis, depending on whether the linearity assumption holds or is violated. If the assumption is met, the analysis can proceed as planned. However, if the assumption is not met, adjustments to the model or alternative methods may be necessary to address the non-linearity appropriately." - [Binomial Logistic Regression](https://www.amstatisticalconsulting.com/binomial-logistic-regression/)


Basically, this equation needs to have a somewhat linear relationship with the independent variable: 

$$g(p) = log\left( \frac{p}{1-p} \right)$$
Where p = probability of 1 (if 1, p. if 0, 1-p)?

```{r}
eDNAxEnvDataSat_SOI_box <- eDNAxEnvDataSat_SOI %>% 
  filter(!is.na(SatPct), !is.na(Present), !is.na(year)) %>% 
  filter(Species %in% c("Calanus pacificus", "Clupea pallasii", "Oncorhynchus tshawytscha")) # For some goddamn reason there's a species or two that REFUSES to box tidwell
joinSpeciesList <- split(eDNAxEnvDataSat_SOI_box, eDNAxEnvDataSat_SOI_box$Species) # Split by species

library(car) # Has a function for the Box-Tidwell procedure
?boxTidwell

# Single test
boxTidwell(Present ~ SatPct, other.x = ~ year, data = joinSpeciesList[[1]]) # other.x = any factors not to be transformed. i had to make year into a factor to make it accept this, not sure what it's doing.

# Loop to test all for linearity with DO
for (i in 1:length(joinSpeciesList)) {
  print(paste(names(joinSpeciesList)[i], sep = " ", "Presence vs Oxygen"))
  print(boxTidwell(Present ~ SatPct, other.x = ~ year, data = joinSpeciesList[[i]]))
}
```

## Calculate regression

```{r}
oxmodels <- lapply(joinSpeciesList, glm, formula = Present ~ SatPct, family = "binomial") # Make the model for each species
lapply(oxmodels, summary) # Print the summary of each model
```

## Looped Binomial Regression

```{r}
# First, filter for more than 50 detections

Detect50Sum <- eDNAxEnvDataSat %>% 
  group_by(Species) %>% 
  summarize(Detections = sum(Present)) %>% 
  arrange(desc(Detections)) %>% 
  filter(Detections > 50 & !is.na(Species)) 
Detect50Species <- Detect50Sum$Species # 37 species

Detect50 <- eDNAxEnvDataSat %>% 
  filter(Species %in% Detect50Species) %>% 
  mutate(year = as.factor(year)) %>% # Clean for boxtidwell
  filter(!is.na(Present) & !is.na(SatPct) & !is.na(year))

Detect50Split <- split(Detect50, Detect50$Species) # Split by species

# Make a function that'll do the binomial regression AND put the results in a dataframe

SpeciesMod <- c()
InterceptP <- c()
SlopeP <- c()
SampleSizes <- c()

#modtest <- glm(Detect50Split[[1]], formula = Present ~ SatPct, family = "binomial")
#modsum <- summary(modtest)
#Coefficients: modsum$coeffifients[1 = intercept/2 = slope, 4]

for (i in 1:length(Detect50Split)) {
  SpName <- names(Detect50Split)[i]
  SpeciesMod <- c(SpeciesMod, SpName)
  
  df <- Detect50Split[[i]]
  
  SampleSize <- sum(df$Present)
  SampleSizes <- c(SampleSizes, SampleSize)
  
  # lapply(joinSpeciesList, glm, formula = Present ~ SatPct, family = "binomial")
  mod <- glm(df, formula = Present ~ SatPct, family = "binomial")
  modsum <- summary(mod)
  
  Intercept <- round(modsum$coefficients[1,4], digits = 5)
  Slope <- round(modsum$coefficients[2,4], digits = 5)
  InterceptP <- c(InterceptP, Intercept)
  SlopeP <- c(SlopeP, Slope)
}

BinomialRegressionResults <- data.frame(Species = SpeciesMod, InterceptP, SlopeP, n = SampleSizes)

SignificantSlopes <- BinomialRegressionResults %>% filter(SlopeP < 0.05)
SignificantIntercepts <- BinomialRegressionResults %>% filter(InterceptP < 0.05)

SigSp <- SignificantSlopes$Species
SigSpDf <- eDNAxEnvDataSat %>% 
  filter(Species %in% SigSp)
```

 [1] "Paracalanus sp. C AC-2013"!!bad      "Chaetoceros socialis"          
 [3] "Lithodesmium intricatum"        "Cyphoderia ampulla"            
 [5] "Bathycoccus prasinos"           "Citharichthys sordidus"        
 [7] "Clytia gregaria"                "Paramoeba branchiphila"        
 [9] "Vannella sp."                   "Parvamoeba rugata"             
[11] "Chrysochromulina sp. NIES-1333" "Aphanomyces astaci"            
[13] "Aureococcus anophagefferens"    "Heterosigma akashiwo"          
[15] "Thecamonas trahens"     

## ALLLLLLL REGRESSIONS
```{r}
# Don't filter for more than 50 detections

eDNAxEnvDataSatClean <- eDNAxEnvDataSat %>% 
  mutate(year = as.factor(year)) %>% # Clean for boxtidwell
  filter(!is.na(Present) & !is.na(SatPct) & !is.na(year)) # Clean for NA values

AllSplit <- split(eDNAxEnvDataSatClean, eDNAxEnvDataSatClean$Species) # Split by species

# Make a function that'll do the binomial regression AND put the results in a dataframe

SpeciesMod2 <- c()
InterceptP2 <- c()
SlopeP2 <- c()
SampleSizes2 <- c()

#modtest <- glm(Detect50Split[[1]], formula = Present ~ SatPct, family = "binomial")
#modsum <- summary(modtest)
#Coefficients: modsum$coeffifients[1 = intercept/2 = slope, 4]

for (i in 1:length(AllSplit)) {
  SpName <- names(AllSplit)[i]
  SpeciesMod2 <- c(SpeciesMod2, SpName)
  
  df <- AllSplit[[i]]
  
  SampleSize <- sum(df$Present)
  SampleSizes2 <- c(SampleSizes2, SampleSize)
  
  mod <- glm(df, formula = Present ~ SatPct, family = "binomial")
  modsum <- summary(mod)
  
  Intercept <- round(modsum$coefficients[1,4], digits = 5)
  Slope <- round(modsum$coefficients[2,4], digits = 5)
  InterceptP2 <- c(InterceptP2, Intercept)
  SlopeP2 <- c(SlopeP2, Slope)
}

ALLBinomialRegressionResults <- data.frame(Species = SpeciesMod2, InterceptP = InterceptP2, SlopeP = SlopeP2, n = SampleSizes2)

ALLSignificantSlopes <- ALLBinomialRegressionResults %>% filter(SlopeP2 < 0.05)
ALLSignificantIntercepts <- ALLBinomialRegressionResults %>% filter(InterceptP2 < 0.05)

ALLSigSp <- ALLSignificantSlopes$Species
ALLSigSpDf <- eDNAxEnvDataSatClean %>% 
  filter(Species %in% ALLSigSp)

system("say Binomial Regression Models Computed")
```

## Box Tidwell Loop
```{r}
SigSpBox <- eDNAxEnvDataSatClean %>%
  filter(!is.na(SatPct)) %>% 
  filter(Species %in% c("Aphanomyces astaci", "Aureococcus anophagefferens", "Thecamonas trahens", "Chaetoceros socialis", "Lithodesmium intricatum", "Cyphoderia ampulla", "Bathycoccus prasinos", "Clytia gregaria", "Paramoeba branchiphila", "Parvamoeba rugata", "Citharichthys sordidus",  "Thermocyclops inversus", "Pennaria disticha", "Clupea pallasii")) # Species I actually want to run this on, NA values removed

fail <- c("Acartia longiremis", "Chrysochromulina sp. NIES-1333", "Stephanopyxis turris", "Vannella sp.", "Heterosigma akashiwo") # Species where box-tidwell failed - they have SatPct NA values but the drop_na should fix that

LinearSplit <- split(SigSpBox, SigSpBox$Species) # Split by species, deleting species that broke the Box-Tidwell test for now

LinearSplit <- lapply(LinearSplit, drop_na, c(SatPct, Present, year))

test <- LinearSplit[[1]]
test2 <- LinearSplit[[2]]

# Do a box Tidwell test on everything that had significant slopes
# This only works on some of the species for some reason?
SpeciesBox <- c()
BoxTidwellP <- c()

for (i in 1:length(LinearSplit)) {
  print(paste(names(LinearSplit)[i], sep = " ", "Presence vs Oxygen"))
  box <- boxTidwell(Present ~ SatPct, 
                    other.x = ~ year, 
                    data = LinearSplit[[i]], 
                    na.action = na.exclude)
  
  x <- round(box$result[3], digits = 5)
  SpeciesBox <- c(SpeciesBox, names(LinearSplit)[i])
  BoxTidwellP <- c(BoxTidwellP, x)
}

BoxTidwellResults <- data.frame(SpeciesBox, BoxTidwellP) %>% arrange(desc(BoxTidwellP))

BoxTidwellSuccess <- BoxTidwellResults %>% filter(BoxTidwellP > 0.05)
```


# Graph presence over DO vs temp

```{r}
Presences <- eDNAxEnvDataSat_SOI %>% 
  filter(Present == T)

ggplot(EnvDataRange, aes(x = temperature, y = SatPct)) +
  geom_point(color = "gray90") +
  geom_point(data = Presences, aes(x = temperature, y = SatPct, color = Species), alpha = 1) +
  theme_bw()
```

# Detection Rates
```{r}
# Add hypoxia column
EnvDataRange <- EnvDataRange %>% 
  mutate(Hypoxia20 = case_when(SatPct < 20 ~ T, .default = F), Hypoxia30 = case_when(SatPct < 30 ~ T, .default = F))
eDNAxEnvDataSat_SOI <- eDNAxEnvDataSat_SOI %>% 
  mutate(Hypoxia20 = case_when(SatPct < 20 ~ T, .default = F), Hypoxia30 = case_when(SatPct < 30 ~ T, .default = F))
eDNAxEnvDataSat <- eDNAxEnvDataSat %>% 
  mutate(Hypoxia20 = case_when(SatPct < 20 ~ T, .default = F), Hypoxia30 = case_when(SatPct < 30 ~ T, .default = F))

# Calculate herring numbers
eDNAxEnvDataSat_SOI %>% 
  filter(Species == "Clupea pallasii", Present == T) %>% 
  group_by(Hypoxia20) %>% 
  summarize(Observations = n()) %>% 
  mutate(Percentage = Observations/sum(Observations))

# Numbers to check with herring
TotalHerring <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii", Present == T) %>% summarize(n = n()))
HypoxicHerring <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii", Present == T, Hypoxia20 == T) %>% summarize(n = n()))
OxicHerring <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii", Present == T, Hypoxia20 == F) %>% summarize(n = n()))

TotalSamples <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii") %>% summarize(n = n()))
HypoxicSamples <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii", Hypoxia20 == T) %>% summarize(n = n()))
OxicSamples <- as.numeric(eDNAxEnvDataSat_SOI %>% filter(Species == "Clupea pallasii", Hypoxia20 == F) %>% summarize(n = n()))
```

```{r}
# Calculating detection rates for all species
hypoxic <- eDNAxEnvDataSat %>% 
  filter(Hypoxia20 == T)

oxic <- eDNAxEnvDataSat %>% 
  filter(Hypoxia20 == F)

# Samps = samples
# Dtct = detections
Hsum <- hypoxic %>% 
  group_by(Species) %>% 
  summarize(Dtct_H = sum(Present), Samps_H = n()) %>% 
  mutate(Pct_Dtct_H = round((Dtct_H/Samps_H)*100, digits = 2))

Osum <- oxic %>% 
  group_by(Species) %>% 
  summarize(Dtct_O = sum(Present), Samps_O = n()) %>% 
  mutate(Pct_Dtct_O = round((Dtct_O/Samps_O)*100, digits = 2))

PercentageComparison <- full_join(Hsum, Osum, by = join_by(Species)) %>% 
  mutate(TotalDtct = Dtct_H+Dtct_O, 
         Samps = Samps_H + Samps_O, 
         Pct_Dtct_Total = round((TotalDtct/Samps)*100, digits = 2), 
         Delta = Pct_Dtct_O - Pct_Dtct_H, # Delta = Oxic - hypoxic detection percentages
         AbsDelta = abs(Delta)) # To sort by

# Filter for large differences between hypoxic and non-hypoxic detection rates
RateChange <- PercentageComparison %>% 
  filter(abs(Delta) > 20)
BigRateChange <- PercentageComparison %>% 
  filter(abs(Delta) > 50)
```

# New SOI list

```{r}
# Acquire the data I want
EllaInterest <- read_csv(here("OCNMS_Project", "Data", "EllaInterestStats.csv")) %>%  # Importing my spreadsheet of species of interest based on all of those tests. Indicates whether the species are significant by detection rate (>30%) and whether they had a significant binomial regression result with oxygen saturation.
  mutate(DeltaSig20 = case_when(abs(DetectionRateDelta) >= 0.2 ~ T, .default = F)) %>% 
  relocate(DeltaSig20, .after = DeltaSig30)
SigBoth20 <- EllaInterest %>% filter(BinomSig == T & DeltaSig20 == T) # Extract species names that are significant based on binomial regression and change in detection rates above and below the hypoxic threshold
SigBoth20 <- SigBoth20$Species
SigBoth <- as.vector(EllaInterest %>% filter(BinomSig == T & DeltaSig30 == T) %>% select(Species))
SigBinom <- as.vector(EllaInterest %>% filter(BinomSig == T) %>% select(Species)) 
SigDelta <- as.vector(EllaInterest %>% filter(DeltaSig30 == T) %>% select(Species)) 

SigBoth20df <- eDNAxEnvDataSatClean %>% 
  filter(Species %in% SigBoth20)

EllaInterestDF <- eDNAxEnvDataSatClean %>% 
  filter(Species %in% EllaInterest$Species)
```

# Graphs (Part 2)

```{r}
source(here("OCNMS_Project", "PresenceEnvGraphFunction.R"))
x <- test(2) # Should print 4 and assign x to 6
presenceGraphVars()

# Test the function
presenceGraph(eDNAxEnvDataSat_SOI, envCond = "SatPct", envCondName = "Oxygen", thresholdLvl = 30, widthpx = 3000, ylab = "Oxygen Saturation (%)")
```

```{r}
# Run the function on my data of interest
presenceGraph(EllaInterestDF, envCond = "SatPct", envCondName = "Oxygen", thresholdLvl = 20, heightpx = 1500, widthpx = 3000, ylab = "Oxygen Saturation (%)")

presenceGraph(EllaInterestDF, envCond = "temperature", envCondName = "Temperature", thresholdLvl = 20, heightpx = 1500, widthpx = 3000, ylab = "Temperature (ºC)")

system("say Presence and absence graphs complete!")
```

# Graph binomial regression

```{r}
binomtest <- EllaInterestDF %>% filter(Species == "Thecamonas trahens") %>% mutate(PresentBin = case_when(Present == T ~ 1, Present == F ~ 0))
ggplot(binomtest, aes(x = SatPct, y = PresentBin)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = F) +
  stat_poly_eq(use_label(c("eq", "R2"))) +
  theme_bw()
```

# Loop to graph binom regresion
```{r}
for (i in 1:length(AllSplit)) {
  df <- AllSplit[[i]] %>% mutate(PresentBin = case_when(Present == T ~ 1, Present == F ~ 0))
  species <- as.character(AllSplit[[i]][1,7])
  system(paste("say", species))
  # print(head(df))
  
  ggplot(df, aes(x = SatPct, y = PresentBin)) +
    geom_point() +
    geom_smooth(method = "glm", method.args = list(family = "binomial"), se = F) +
    stat_poly_eq(use_label(c("eq", "R2"))) +
    theme_bw() +
    labs(title = species)
  
    ggsave(filename = here("OCNMS_Project", "Plots",
                           "Binomial_Regressions", 
                           paste(species, sep = ".", "png")), 
           width = 2300, 
           height = 2000, 
           units = "px")
}

system("say Binomial Regression Graphs Complete!")
```


S Xu#, M Chen#, T Feng, L Zhan, L Zhou, G Yu*. Use ggbreak to effectively utilize plotting space to deal with large datasets and outliers. Frontiers in Genetics. 2021, 12:774846. doi: 10.3389/fgene.2021.774846

```{r}
system("say Script Complete! Beep Beep!")
```

To kill the talking enter in terminal `while true; do pkill -9 say; done`


