---
title: "Environmental Data In Dates of Interest"
author: "Eleanor (Ella) Crotty"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    toc: TRUE
urlcolor: blue
---

```{r Package Imports, message = F, warning = F}
# Warnings and startup messages suppressed
library(tidyverse)
library(patchwork)
library(scales)
library(ggrepel)
library(readxl)
library(here)
library(gghighlight)
```

# Manually enter dates of interest (old)

```{r, eval = F, show = F}
# Define dates of interest (old, manual)

BottleDates_Manual <- as.POSIXct(c("2021-08-25", "2022-06-21", "2022-08-22", "2022-09-20", "2021-10-08", "2022-07-19", "2023-06-15", "2023-07-26")) # Times from these are available
# seq() will take POSIXct objects, and by can be specified in seconds. Every 36 hours * * 60 minutes * 60 seconds = sampling every 129600 seconds
# Needs to be POSIXct for graphing

by1 <- 129600

PPS_Dates_Manual <- c(
  seq(from = as.POSIXct("2022-08-23"), by = by1, to = as.POSIXct("2022-09-18")), 
  seq(from = as.POSIXct("2021-08-26"), by = by1, to = as.POSIXct("2021-09-29")), 
  seq(from = as.POSIXct("2022-06-23"), by = by1, to = as.POSIXct("2022-07-19")), 
  seq(from = as.POSIXct("2023-06-16"), by = by1, to = as.POSIXct("2023-07-20"))
  )

Source <- "Bottle_DNA"
BottleDatesDF <- data.frame(Source, BottleDates_Manual) %>% rename(Date = BottleDates_Manual)
Source <- "PPS_DNA"
PPS_DatesDF <- data.frame(Source, PPS_Dates_Manual) %>% rename(Date = PPS_Dates_Manual)
DatesOfInterest_Manual <- rbind(BottleDatesDF, PPS_DatesDF)
# This can now be used for filtering joins :)

write.csv(DatesOfInterest_Manual, file = here("OCNMS_Project", "Data", "SamplingDates_Manual.csv"))
```

# Import sample dates

```{r}
eDNA_Meta <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_eDNA_SampleData.csv"))
eDNA_dates <- eDNA_Meta[4] # why does this open the file
colnames(eDNA_dates) <- "date"
eDNA_times <- eDNA_Meta[13]
colnames(eDNA_times) <- "Time"
# eDNA_dates$Date <- as.POSIXct(eDNA_dates$Date, tryFormats = c("%m/%d/%Y")) # Needs to be POSIXct for graphing

eDNA_positions <- eDNA_Meta[5]
colnames(eDNA_positions) <- "Position"
# Bind eDNA_dates to eDNA_positions, mutate(case_when contains "Port" = PPS, ELSE source = Bottle)

# 24 hr time = %H%M
DatesOfInterest2 <- data.frame(eDNA_dates, eDNA_times, eDNA_positions) %>% 
  mutate(Source = case_when(str_detect(Position, "Port") ~ "PPS_DNA", .default = "Bottle_DNA")) %>% 
  mutate(date = paste(date, Time)) %>% 
  mutate(Date = as.POSIXct(date, tryFormats = c("%m/%d/%Y %H:%M"), tz = "America/Los_Angeles")) %>%  # it keeps the same time - not sure this is what i'm going for
  mutate(Date = with_tz(Date, tzone = "UTC")) %>%  # Same instant in time, displays different time and timezone, now I can mash it with UTC stuff
  select(Date, Source)

write.csv(DatesOfInterest2, file = here("OCNMS_Project", "Data", "SamplingDates2.csv"))
```

# Import environmental data

```{r}
MooringCTD_Data <- read.csv(here("OCNMS_Hypoxia", "Outputs", "EnvironmentalDataset1.csv")) %>%  # Import mooring and CTD environmental data from 2021-2023 + 30-45 meters deep
  select(-X)
MooringCTD_Data$date <- as.POSIXct(MooringCTD_Data$date, tz = "UTC") # time stays the same. good

NEMO_Data <- read.csv(here("OCNMS_Hypoxia", "Outputs", "NEMO_Data_Clean.csv")) %>%  # checked date - it does seem to be the same POSIXct time as before export
  select(-X)
NEMO_Data$date <- as.POSIXct(NEMO_Data$date, tz = "UTC") # time stays the same. good
NEMO_Data$source <- "NEMO" # Prep for combination

# Trying to get more efficient at adding all the NA columns for the combination
NEMO_Data <- NEMO_Data %>% 
  mutate(salinity = NA, potential_density = NA, pres = NA, cond = NA, sampleID = NA) %>% 
  select("source", "date", "year", "temperature", "DO", "salinity", "potential_density", "pres", "cond", "sampleID")

Moor_CTD_NEMO_Data <- rbind(MooringCTD_Data, NEMO_Data)
```

# Plot manual sample dates (old)

```{r, eval = F, show = F}
# Plot sampling (manual) vs mooring data

ggplot(MooringCTD_Data, aes(x = date, y = temperature, color = source)) +
  geom_point() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest_Manual, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "black", "dodgerblue2", "purple1"))

ggplot(MooringCTD_Data, aes(x = date, y = DO, color = source, alpha = source)) +
  geom_point() +
  scale_alpha_manual(values = c(1,1)) + # doesn't currently do anything
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest_Manual, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia", "Plots", "OxygenOverDeploymentWindows.png"))

ggplot(MooringCTD_Data %>% filter(source == "CTD"), aes(x = date, y = DO, color = source)) +
  geom_point() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest_Manual, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia", "Plots", "OxygenOverDeploymentWindowsCTD.png"))
# The CTD and Bottle_DNA dots should overlap!
```

# Plot imported samples

```{r}
# Plot sampling (imported from Shannon's spreadsheet) vs mooring data

ggplot(MooringCTD_Data, aes(x = date, y = temperature, color = source)) +
  geom_point() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest2, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "black", "dodgerblue2", "purple1"))

ggplot(MooringCTD_Data, aes(x = date, y = DO, color = source, alpha = source)) +
  geom_point() +
  scale_alpha_manual(values = c(1,1)) + # doesn't currently do anything
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest2, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia", "Plots", "OxygenOverDeploymentWindows2.png"))

ggplot(MooringCTD_Data %>% filter(source == "CTD"), aes(x = date, y = DO, color = source)) +
  geom_point() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  geom_vline(data = DatesOfInterest2, aes(xintercept = Date, color = Source)) +
  scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia", "Plots", "OxygenOverDeploymentWindowsCTD2.png"))
# The CTD and Bottle_DNA dots should overlap!
```

## work in progress, trying to make CTD show better
ggplot(DatesOfInterest, aes(xintercept = Date, color = Source)) +
  geom_vline() +
  geom_point(MooringCTD_Data, aes(x = date, y = DO, color = source, alpha = source), inherit.aes = F) +
  scale_alpha_manual(values = c(1,0.01)) +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia2", "Plots", "OxygenOverDeploymentWindows.png"))

# Troubleshooting

```{r}
# Alright, where's the problem
metatrouble <- DatesOfInterest2 %>% filter(Source == "Bottle_DNA") %>% group_by(Date) %>% summarize(Source = min(Source)) # Dates returned by spreadsheet import
publishtrouble <- MooringCTD_Data %>% filter(source == "CTD") %>% select(source, date) # Dates from CTD data from drive (from Sam's pipeline + my cleaning)

trs <- read.csv(here("OCNMS_eDNA", "Data", "SampleDateRecords.csv")) # trs for troubleshoot
DateUTC <- trs[8]
LocalDate <- as.vector(trs[9])
Position <- as.vector(trs[12])
StartLocal <- as.vector(trs[23])
StartUTC <- as.vector(trs[25])
EndLocal <- as.vector(trs[24])
EndUTC <- as.vector(trs[26])

trs2 <- data.frame(DateUTC, LocalDate, Position, StartLocal, StartUTC, EndLocal, EndUTC)
```

# Plot NEMO + imported samples

```{r}
# Plot sampling (imported from Shannon's spreadsheet) vs mooring data

ggplot(Moor_CTD_NEMO_Data, aes(x = date, y = temperature, color = source, alpha = source)) +
  geom_point(shape = 1, size = 0.5) +
  scale_alpha_manual(values = c(1,0.3,0.5)) +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  #geom_vline(data = DatesOfInterest2, aes(xintercept = Date, color = Source)) +
  #scale_color_manual(values = c("black", "black", "dodgerblue2", "orange", "purple1"))
  scale_color_manual(values = c("red", "dodgerblue2", "purple1"))
ggsave(here("OCNMS_Hypoxia", "Plots", "TempOverDeploymentWindowsNEMO.png"))

ggplot(Moor_CTD_NEMO_Data, aes(x = date, y = DO, color = source, alpha = source)) +
  geom_point(shape = 1, size = 0.5) +
  scale_alpha_manual(values = c(1,0.5,0.5)) + 
  scale_x_datetime(date_breaks = "1 month", date_labels = "%m-%y") +
  theme_bw() +
  facet_wrap(facets = vars(year), scales = "free_x", ncol = 1) +
  # geom_vline(data = DatesOfInterest2, aes(xintercept = Date, color = Source)) +
  # scale_color_manual(values = c("black", "firebrick2", "dodgerblue2", "orange", "purple1")) +
  scale_color_manual(values = c("red", "dodgerblue2", "purple1")) +
  geom_hline(aes(yintercept = 2), color = "black", linetype = "dashed") +
  labs(caption = "Dotted line = hypoxic threshold of 2 mg/L dissolved oxygen")
ggsave(here("OCNMS_Hypoxia", "Plots", "OxygenOverDeploymentWindowsNEMO.png"))
```
