---
title: "eDNA Exploration (Take 2)"
author: "Eleanor (Ella) Crotty"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    toc: TRUE
urlcolor: blue
---

```{r Package Imports, message = F, warning = F}
# Warnings and startup messages suppressed
library(tidyverse)
library(patchwork)
library(scales)
library(ggrepel)
library(readxl)
library(here)
```

# Import eDNA combined data

```{r}
# Remove non-TH042 samples
# Remove Chris Paight samples
fishReads <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_MiFish_Data.csv"))
fishReads <- fishReads %>% 
  mutate(Marker = "MiFish") %>% 
  filter(Cast_No. == "TH042") 
# No more Chris. yay.

OCNMS_metadata <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_eDNA_sampleData_full.csv"))

coiReads <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_COI_Data.csv"))
coiReads <- coiReads %>% 
  mutate(Marker = "Uni_CO1", Barcode_mod = NA) %>% # Barcode_mod just needs to be there to match fish.
  relocate(Barcode_mod, .after = Barcode.x) %>% 
  filter(Cast_No. == "TH042") 
```

# Make allReads

```{r}
allReads <- rbind(fishReads, coiReads) %>% filter(ESV != "ESV_225890")
write_csv(allReads, here("OCNMS_eDNA", "Outputs", "FishPlusCOI_Reads.csv")) # Save all the reads for reference
```

## Deal with PCR replicates

```{r}
PCR_reps_combine <- function(df) {
  # print(head(df))
  
  # Separate out Sample_Name into three informational columns
  df_out <- df %>% 
    separate(Sample_Name, 
             into=c("E_no", "Cruise1", "PCR_Rep"), 
             remove=F, 
             fill = "right") %>% # sep is a regular expression which is annoying, but the default recognizes any non-alphanumeric characters so the default works here
    mutate(.)
  
  # Define which columns to group by
  id_cols1 <- colnames(df_out) # Pull out column names
  id_cols1 <- id_cols1[! id_cols1 %in% c("X", "SampleId","Sample_Name", "PCR_Rep", "pctMatch", "JV_Sample_Name", "Technical_Replicate", "nReads", "Barcode.y")] # Remove SampleId and Sample_Name, as well as other things that differ by PCR replicate
  print(id_cols1)
  
  df_out <- df_out %>% 
    group_by_at(id_cols1) %>% 
    # dplyr::mutate(SampleId = dplyr::first(SampleId)) %>% # Just ignoring sampleid for now and using E-no as unique
    summarize(nReads = mean(nReads)) %>% 
    relocate(nReads, .after = PI)
    
  df_out
}

df_out1 <- PCR_reps_combine(allReads) # Spot check on E1032 worked

allReads_PCRcomb <- PCR_reps_combine(allReads)
```


## Group by species
## Export AllReads

```{r}
# Replace with allReads_PCRcomb when done

id_cols <- colnames(allReads_PCRcomb) # Pull out column names
id_cols <- id_cols[! id_cols %in% c("X", "ESV", "sequence", "nReads")] # Remove ESV + nreads

id_cols2 <- colnames(fishReads) # Pull out column names
id_cols <- id_cols[! id_cols %in% c("X", "ESV", "sequence", "nReads", "pctMatch")] # Remove ESV + nreads + pctMatch

allReads_species <- allReads_PCRcomb %>% 
  group_by_at(id_cols) %>% # group_by_at can take a vector
  summarize(TotalnReads = sum(nReads)) %>% # Removed , avgpctMatch = mean(pctMatch) because I had to remove pctMatch to combine PCR replicates
  relocate(TotalnReads, .after = PI)

write_csv(allReads_species, here("OCNMS_eDNA", "Outputs", "AllReads_Species.csv"))

fishReads_species <- fishReads %>% 
  group_by_at(id_cols2) %>% # group_by_at can take a vector
  summarize(TotalnReads = sum(nReads)) %>% 
  relocate(TotalnReads, .after = PI)

coiReads_species <- coiReads %>% 
  group_by_at(id_cols2) %>% # group_by_at can take a vector
  summarize(TotalnReads = sum(nReads)) %>% 
  relocate(TotalnReads, .after = PI)
```

# Grab ESVs of interesting species

See eDNAxpO2.Rmd for why these species are interesting. This is just to generate reference ESVs for a BLAST search.

```{r}
EllaInterest <- read_csv(here("OCNMS_Project", "Data", "EllaInterestStats.csv")) %>%  # Importing my spreadsheet of species of interest based on all of those tests. Indicates whether the species are significant by detection rate (>30%) and whether they had a significant binomial regression result with oxygen saturation.
  mutate(DeltaSig20 = case_when(abs(DetectionRateDelta) >= 0.2 ~ T, .default = F)) %>% 
  relocate(DeltaSig20, .after = DeltaSig30)

EllaInterestDF <- allReads %>% 
  filter(Species %in% EllaInterest$Species)

ESVfinder <- EllaInterestDF %>% 
  group_by(ESV, sequence, Species) %>% 
  summarize(Samples = n(), AvgPctMatch = mean(pctMatch), Totalreads = sum(nReads)) %>% 
  arrange(Species)

litESVfinder <- ESVfinder %>% 
  filter(Species %in% c("Acanthamoeba comandoni", "Aphanomyces astaci", "Aureococcus anophagefferens", "Bathycoccus prasinos", "Citharichthys sordidus", "Clupea pallasii", "Clytia gregaria", "Heterosigma akashiwo", "Lithodesmium intricatum", "Paracalanus sp. C AC-2013", "Pennaria disticha", "Poterioochromonas stipitata", "Pseudocalanus acuspes", "Stephanopyxis turris", "Thecamonas trahens", "Thermocyclops inversus"))
```


# Import species of interest

```{r}
SOI_df <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_SpeciesOfInterest.csv")) 
SOI_invert_df <- read.csv(here("OCNMS_eDNA", "Data", "OCNMS_CopepodsAndKrillofInterest.csv")) 
genera <- c("Pseudo-nitzchia", "Pyrosoma", "Pyrosomella", "Sargassum", "Undaria", "Acrozoanthus", "Isarus", "Zoanthus")
families <- c("Ammodytidae", "Acipenseridae")
SpeciesOfInterest <- c(SOI_df$Taxon, SOI_invert_df$Taxon)
```

# SOI Functions

```{r}
# Summaries

SOI_fn <- function(df) {
  marker <- df$Marker[1] # This won't work on allReads
  
  s <- df %>% 
    group_by(Species) %>% 
    summarize(Count = n(), MeanPctMatch = mean(pctMatch), Genus = NA, Family = NA) %>% 
    filter(Species %in% SpeciesOfInterest) %>% 
    select(Family, Genus, Species, Count, MeanPctMatch) %>% 
    mutate(Marker = marker)
  
  g <- df %>% 
    group_by(Genus) %>% 
    summarize(Count = n(), MeanPctMatch = mean(pctMatch), Species = NA, Family = NA) %>% 
    filter(Genus %in% genera) %>% 
    select(Family, Genus, Species, Count, MeanPctMatch) %>% 
    mutate(Marker = marker)
  
  f <- df %>% 
    group_by(Family) %>% 
    summarize(Count = n(), MeanPctMatch = mean(pctMatch), Genus = NA, Species = NA) %>% 
    filter(Family %in% families) %>% 
    select(Family, Genus, Species, Count, MeanPctMatch) %>% 
    mutate(Marker = marker)
  
  rbind(s,g,f)
}

MiFish_SOI_id_summary <- SOI_fn(fishReads)
COI_SOI_id_summary <- SOI_fn(coiReads)
SOI_id_summary <- rbind(MiFish_SOI_id_summary, COI_SOI_id_summary)
```

```{r}
# Now to make a df of all the detections (not a summary per species & marker)
SOI_full_fn <- function(df) {
  marker <- df$Marker[1]
  s <- df %>% 
    filter(Species %in% SpeciesOfInterest | Genus %in% genera | Family %in% families) %>% 
    mutate(Marker = marker)
  s
}

MiFish_SOI_IDs <- SOI_full_fn(fishReads)
COI_SOI_IDs <- SOI_full_fn(coiReads)
SOI_IDs <- rbind(MiFish_SOI_IDs, COI_SOI_IDs) %>% 
  group_by(Family, Genus, Species)
```
Old code:
```
# Group by species, not ESV
SOI_ID_Species <- SOI_IDs %>% 
  group_by(Family, Genus, Species, JV_Sample_Name, Barcode.x, Barcode_mod, SampleId, Biological_Replicate, Technical_Replicate, Cruise_ID_short, Cast_No., Rosette_position, Field_collection_method, Preservation, Station, Depth_m, Field_Notes, Date_UTC, Date_local, Lat_dec, Lon_dec, Run, Client, Replicate, Amplicon, Barcode.y, Marker) %>% 
  summarize(TotalnReads = sum(nReads), avgpctMatch = mean(pctMatch))
```

I named this SOI_ID_Speciestest in the next script to see if it matched when I ran theortically similar code here. It matches!

  
```{r}
id_cols <- colnames(SOI_IDs) # Pull out column names
id_cols <- id_cols[-c(1,2,3,11,18)] # Remove ESV + pctMatch + nreads

SOI_ID_Species <- SOI_IDs %>% 
  group_by_at(id_cols) %>% # group_by_at can take a vector
  summarize(TotalnReads = sum(nReads), avgpctMatch = mean(pctMatch))
```


# Export SOI

```{r}
write_csv(MiFish_SOI_IDs, here("OCNMS_eDNA", "Data", "MiFish_SOI_IDs.csv"))
write_csv(COI_SOI_IDs, here("OCNMS_eDNA", "Data", "COI_SOI_IDs.csv"))
write_csv(SOI_IDs, here("OCNMS_eDNA", "Data", "All_SOI_IDs.csv"))
write_csv(SOI_ID_Species, here("OCNMS_eDNA", "Data", "All_SOI_IDs_Species.csv"))
```


```{r}
SOI_IDsyr <- SOI_IDs %>% 
  mutate(Date_UTC = as.POSIXct(Date_UTC, tz = "UTC"), year = as.factor(year(as.POSIXct(Date_UTC, tz = "UTC"))))
# Take allReads and start poking at dates and stuff!
```

```{r}
ggplot(SOI_IDsyr, aes(x = Species)) +
  geom_bar() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
SOI_IDs %>% 
  group_by(Family, Genus, Species) %>% 
  summarize(n = n()) %>% 
  arrange(n)

SOI_IDs %>% 
  group_by(Family, Genus, Species) %>% 
  summarize(n = n()) %>% 
  arrange(n) %>% 
  filter(n > 10)
```

# Calculate eDNA index

Warning: This uses the data with PCR replicates NOT combined (except allReads_species, which does have them combined)

```{r}
index1dummy <- allReads %>%
  dplyr::group_by(SampleId) %>% 
  mutate(Tot = sum(nReads),
      	Row.sums = nReads / Tot) %>% # calculate proportions - 0 reads/0 total = NaN, need to replace with 0 to make max() work
  relocate(c(Tot, Row.sums), .after = SampleId)
```

## There's a bunch of sample with no reads of any species

Just gonna filter those out for now

```{r}
problemchildren <- index1dummy %>% 
  filter(Tot == 0)
problemchildrencount <- problemchildren %>% 
  group_by(SampleId) %>% 
  summarize(n = n())
problemchildrenID <- problemchildrencount$SampleId
problem_meta <- OCNMS_metadata %>% 
  filter(SampleId %in% problemchildrenID)
problem_full <- allReads %>% 
  filter(SampleId %in% problemchildrenID)
write.csv(problem_meta, here("OCNMS_eDNA", "Data", "Flag", "DetectionProblem_Metadata.csv"))
write.csv(problem_full, here("OCNMS_eDNA", "Data", "Flag", "DetectionProblem_Full.csv"))
```

## MiFish

### By ESV

```{r}
MF_index1 <- fishReads %>%
  filter(!(SampleId %in% problemchildrenID)) %>% 
  dplyr::group_by(SampleId) %>% 
  mutate(Tot = sum(nReads),
      	Row.sums = nReads / Tot) %>% # calculate proportions - 0 reads/0 total = NaN, need to replace with 0 to make max() work
  relocate(c(nReads, Tot, Row.sums), .after = SampleId)

MF_eDNA_index_ESV <- MF_index1 %>% 
  dplyr::group_by(ESV) %>%
  mutate(Colmax = max(Row.sums), Normalized.reads = Row.sums / Colmax) %>%  #transforms raw number of reads to eDNA index
  mutate(Normalized.reads = case_when(Normalized.reads == "NaN" ~ 0, 
                                      .default = Normalized.reads)) %>% 
  relocate(c(Colmax, Normalized.reads), .after = Row.sums) 
```

### By species
```{r}
# Index by species 

MF_index1_sp <- fishReads_species %>%
  filter(!(SampleId %in% problemchildrenID)) %>% 
  dplyr::group_by(SampleId) %>% 
  mutate(Tot = sum(TotalnReads),
      	Row.sums = TotalnReads / Tot) %>% # calculate proportions - 0 reads/0 total = NaN, need to replace with 0 to make max() work
  relocate(c(TotalnReads, Tot, Row.sums), .after = SampleId)

MF_eDNA_index_Species <- MF_index1_sp %>% 
  dplyr::group_by(Species) %>%
  mutate(Colmax = max(Row.sums), Normalized.reads = Row.sums / Colmax) %>%  #transforms raw number of reads to eDNA index
  mutate(Normalized.reads = case_when(Normalized.reads == "NaN" ~ 0, 
                                      .default = Normalized.reads)) %>% 
  relocate(c(Colmax, Normalized.reads), .after = Row.sums) 
```

## CO1

### By ESV

```{r}
C_index1 <- coiReads %>%
  filter(!(SampleId %in% problemchildrenID)) %>% 
  dplyr::group_by(SampleId) %>% 
  mutate(Tot = sum(nReads),
      	Row.sums = nReads / Tot) %>% # calculate proportions - 0 reads/0 total = NaN, need to replace with 0 to make max() work
  relocate(c(nReads, Tot, Row.sums), .after = SampleId)

C_eDNA_index_ESV <- C_index1 %>% 
  dplyr::group_by(ESV) %>%
  mutate(Colmax = max(Row.sums), Normalized.reads = Row.sums / Colmax) %>%  #transforms raw number of reads to eDNA index
  mutate(Normalized.reads = case_when(Normalized.reads == "NaN" ~ 0, 
                                      .default = Normalized.reads)) %>% 
  relocate(c(Colmax, Normalized.reads), .after = Row.sums) 
```

### By species (for export)
```{r}
# Index by species 

C_index1_sp <- coiReads_species %>%
  filter(!(SampleId %in% problemchildrenID)) %>% 
  dplyr::group_by(SampleId) %>% 
  mutate(Tot = sum(TotalnReads),
      	Row.sums = TotalnReads / Tot) %>% # calculate proportions - 0 reads/0 total = NaN, need to replace with 0 to make max() work
  relocate(c(TotalnReads, Tot, Row.sums), .after = SampleId)

C_eDNA_index_Species <- C_index1_sp %>% 
  dplyr::group_by(Species) %>%
  mutate(Colmax = max(Row.sums), Normalized.reads = Row.sums / Colmax) %>%  #transforms raw number of reads to eDNA index
  mutate(Normalized.reads = case_when(Normalized.reads == "NaN" ~ 0, 
                                      .default = Normalized.reads)) %>% 
  relocate(c(Colmax, Normalized.reads), .after = Row.sums) 
```

# Export eDNA index
```{r}
write_csv(C_eDNA_index_Species, here("OCNMS_Project", "Data", "COI_eDNA_index.csv"))
write_csv(MF_eDNA_index_Species, here("OCNMS_Project", "Data", "MiFish_eDNA_index.csv"))
```

