---
title: "Manuscript Plots 2026"
author: "Eleanor (Ella) Crotty"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document:
    toc: TRUE
urlcolor: blue
---

```{r Package Imports, message = F, warning = F}
# Warnings and startup messages suppressed
library(tidyverse)
library(patchwork)
library(scales)
library(ggrepel)
library(ggpmisc)
library(here)
library(gsw)
library(R.matlab)
library(ggbreak)
library(car) # Has a function for the Box-Tidwell procedure
library(reshape2) # To make the ASV table into something that can be combined with environmental data
```

# Import data 

```{r}
Sp_ASV_18Sv4 <- read_csv(here("Manuscript_Plots", "OCNMS_final_data", "18Sv4", "Decontam_final_taxonomy_table_pr2.csv")) # 18Sv4 data cleanedd by Sam
envData <- read_csv(here("Manuscript_Plots", "OCNMS_final_data", "18Sv4", "Decontam_final_metadata.csv")) %>% # metadata cleaned by Sam
  mutate(year = as.factor(year(eventDate))) %>%
    relocate(year, .after = eventDate)
ASV_table <- read_csv(here("Manuscript_Plots", "OCNMS_final_data", "18Sv4", "Decontam_final_asv_table.csv"))
```


# Join the data

Goal: 

| sample name | lat | lon | species | detected t/f | 
|-------------|-----|-----|---------|--------------|
| s1          | 1.  | 2.  | fish.   | T            |

```{r Data join}
# The metadata is associated with samples so I think it doesn't need rounding

# I need to combine observations with environmental data
# ASV_table has sample names as columns
# Reads_Species identifies sequences as species

Rds_Txa_18Sv4_pre <- full_join(Sp_ASV_18Sv4, ASV_table)

transpose_test <- Rds_Txa_18Sv4_pre[1:5,]

transposed_test <- melt(transpose_test, id.vars = c("seq_id", "iter", "ASV", "Taxon", "Confidence"), variable.name = "Sample", value.name = "Reads")

# okay, that came out right, here we go. pray for my RAM or whatever.

Rds_Txa_18Sv4 <- melt(Rds_Txa_18Sv4_pre, id.vars = c("seq_id", "iter", "ASV", "Taxon", "Confidence"), variable.name = "Sample", value.name = "Reads")
system("say Transpose complete")

eDNAxEnvData <- full_join(Rds_Txa_18Sv4, envData, by = join_by(Sample == samp_name))
system("say Data Join 1 complete")
```

# Export the combined data

```{r}
# Take a looksee to see which dissolved things are actually collected
# View(eDNAxEnvData[110:150])
smaller <- eDNAxEnvData %>% select(Taxon, iter, Reads, decimalLongitude, decimalLatitude, eventDate, year, samp_weather, temp, chlorophyll, light_intensity, ph, ph_meth, salinity, tidal_stage, turbidity, water_current, wind_speed, percent_oxygen_sat, rosette_position, expedition_id, diss_oxygen, diss_oxygen_unit)

system("say Data Join 2 Complete")

# Export the joined data
write_csv(eDNAxEnvData, here("Manuscript_Plots", "OCNMS_final_Data", "18Sv4_eDNAxEnvData.csv"))
```

# Sampling Window Plots

```{r}
md <- envData

unique(md$expedition_id)

unique(md$samp_collect_method)

md_pps <- md %>% 
  filter(samp_collect_device == "McLane Particle and Phytoplankton Sampler")

sampledatelist <- with_tz(as.POSIXct(md_pps$eventDate, 
                             tz = "America/Los_Angeles",
                             tryFormats = c("%m/%d/%Y")))
sampledatelist <- as.data.frame(sampledatelist) %>%
  mutate(year = year(sampledatelist)) # this is necessary for faceting + individual sample plotting to work together

md_ctd <- md %>% # This is actually the metadata for CTD and the other type of bottle we used when the CTD was broken
  filter(samp_collect_device != "McLane Particle and Phytoplankton Sampler")

ctddatelist <- with_tz(as.POSIXct(md_ctd$eventDate, 
                             tz = "America/Los_Angeles",
                             tryFormats = c("%m/%d/%Y")))
ctddatelist <- as.data.frame(ctddatelist) %>%
  mutate(year = year(ctddatelist))
```

## Aesthetics (vars)
```{r}
widthpx <- 4000 # Width for export (pixels)
heightpx <- 2000 # Height for export (pixels)

widthpx_stacked <- 2000 # Width for export (pixels) for the stacked plots
heightpx_stacked <- 2000 # Height for export (pixels) for the stacked plots

# Rectangle windows aesthetics
width = 1
linecolor = "black"
fillcolor = "navy"

# CTD line or dot aesthetics
ctdcolor = "firebrick4"
# Add CTD as either point or line - uncomment whatever one you want
CTD_add <- geom_point(inherit.aes = F, data = ctddatelist, aes(x = ctddatelist, y = 0), color = ctdcolor)
#CTD_add <- geom_vline(inherit.aes = F, data = ctddatelist, aes(xintercept = ctddatelist), color = ctdcolor)
# Inherit_aes needs to be false, otherwise the main data will override it and cause an error

# Adding text labels
labely <- 0
labelsize <- 3
```

```{r}
# Find start and end dates for expeditions
stend <- envData %>% 
  select(eventDate, expedition_id) %>% 
  filter(year(eventDate) != 2019) %>% # it's two days and doesn't even graph.
  group_by(expedition_id) %>% 
  summarize(start = min(eventDate), end = max(eventDate)) %>%
  mutate(year = year(start),
         x_min = case_when(year == 2021 ~ as.POSIXct("2021-05-01"),
                          year == 2022 ~ as.POSIXct("2022-05-01"),
                          year == 2023 ~ as.POSIXct("2023-05-01")),
         x_max = case_when(year == 2021 ~ as.POSIXct("2021-11-01"),
                          year == 2022 ~ as.POSIXct("2022-11-01"),
                          year == 2023 ~ as.POSIXct("2023-11-01"))) # adds something to graph blank to force full field season to show in facet_wrap 
#scale_x_break(as.POSIXct(c("2021-11-01", "2022-05-01"))) +
  #scale_x_break(as.POSIXct(c("2022-11-01", "2023-05-01"))) 
stend
stend_t <- as.data.frame(t(stend %>% select(start,end)))
stend_t
```

## Sampling Window Graph 1
```{r}
# Graph 1: All in one graph

winplot <- ggplot(data = stend, 
       mapping = aes(x = start, y = 0, xend = end, yend = 0)) +
  theme_bw() +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(size = 12), 
        strip.background = element_rect(fill = "gray95"), # facet label background
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank(),
        axis.text.y = element_text(color = "white")) +
  geom_segment(size = 200, 
       colour = fillcolor,
       alpha = 0.2) +
  # scale break is breaking coord_cartesian(lim) https://github.com/YuLab-SMU/ggbreak/issues/59 put cartesian before
  scale_x_datetime(breaks = "month", date_labels = "%b-%y", limits = as.POSIXct(c("2021-05-01", "2023-11-01"))) +
  #coord_cartesian(ylim = c(-0.5, 0.5)) + # Zooms in the plot without deleting rectangle
  ylim(c(-0.01, 0.01)) +
  # clip = "off" allows data points to be drawn in the margins
  # expand = 0 doesn't add an expansion factor to the limits to keep the data off the axes
  ggtitle("PPS Deployment Windows") +
  xlab("Date") +
  ylab("") +
  scale_x_break(as.POSIXct(c("2021-11-01", "2022-05-01"))) +
  scale_x_break(as.POSIXct(c("2022-11-01", "2023-05-01"))) 

ggsave(plot = winplot,
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesSimple.png"), 
           width = widthpx, 
           height = heightpx, 
           units = "px")
```



## Stacked Years

```{r}
# Graph 2: Stacked Years
# https://stackoverflow.com/questions/42588238/setting-individual-y-axis-limits-with-facet-wrap-not-with-scales-free-y
winplot_stk <- ggplot(data = stend, 
       mapping = aes(x = start, y = 0, xend = end, yend = 0)) +
  theme_bw() +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(size = 12), 
        strip.background = element_rect(fill = "gray95"), # facet label background
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank(),
        axis.text.y = element_text(color = "white")) +
  geom_segment(size = 200, # main plot
       colour = fillcolor,
       alpha = 0.2) +
    # scale break is breaking coord_cartesian(lim)     https://github.com/YuLab-SMU/ggbreak/issues/59 put cartesian before
  scale_x_datetime(breaks = "month", date_labels = "%b-%y") + #, limits = as.POSIXct(c("2021-05-01", "2023-11-01"
  #coord_cartesian(ylim = c(-0.5, 0.5)) + # Zooms in the plot without deleting rectangle
  ylim(c(-0.01, 0.01)) +
  geom_blank(aes(x = x_min)) + # Adds the x_min column to force the x-axis to include the whole field season - it's just some invisible stuff in May and November to make the stacked plot look nice
  geom_blank(aes(x = x_max)) + 
  facet_wrap(vars(year), scales = "free_x", ncol = 1) +
  # clip = "off" allows data points to be drawn in the margins
  # expand = 0 doesn't add an expansion factor to the limits to keep the data off the axes
  ggtitle("PPS Deployment Windows") +
  xlab("Date") +
  ylab("") 

winplot_stk

ggsave(plot = winplot_stk,
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesStacked.png"), 
           width = widthpx_stacked, 
           height = heightpx_stacked, 
           units = "px")
```

## Each Sample Marked

```{r}
# Graph 3: Each sample marked
winplt_mrk <- ggplot() +
  theme_bw() +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(size = 12), 
        strip.background = element_rect(fill = "gray95"), # facet label background
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank(),
        axis.text.y = element_text(color = "white")) +
  geom_segment(data = stend, # data is down here now to avoid conflicts with geom_vline
       mapping = aes(x = start, y = 0, xend = end, yend = 0),
       size = 200, 
       colour = fillcolor,
       alpha = 0.2) +
  geom_vline(inherit.aes = F, 
             data = sampledatelist,
             mapping = aes(xintercept = sampledatelist), 
             color = "black") + # add each individual sample
  # scale break is breaking coord_cartesian(lim) https://github.com/YuLab-SMU/ggbreak/issues/59 put cartesian before
  scale_x_datetime(breaks = "month", date_labels = "%b-%y", limits = as.POSIXct(c("2021-05-01", "2023-11-01"))) +
  #coord_cartesian(ylim = c(-0.5, 0.5)) + # Zooms in the plot without deleting rectangle
  ylim(c(-0.01, 0.01)) +
  # clip = "off" allows data points to be drawn in the margins
  # expand = 0 doesn't add an expansion factor to the limits to keep the data off the axes
  ggtitle("PPS Deployment Windows") +
  xlab("Date") +
  ylab("") +
  scale_x_break(as.POSIXct(c("2021-11-01", "2022-05-01"))) +
  scale_x_break(as.POSIXct(c("2022-11-01", "2023-05-01"))) 

winplt_mrk

ggsave(plot = winplt_mrk,
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesLines.png"), 
           width = widthpx, 
           height = heightpx, 
           units = "px")
```


### Stacked Years

```{r}
# Graph 4: Each Sample Marked + Stacked Years

# https://stackoverflow.com/questions/42588238/setting-individual-y-axis-limits-with-facet-wrap-not-with-scales-free-y
winplt_stk_mrk <- ggplot(data = stend) +
  theme_bw() +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(size = 12), 
        strip.background = element_rect(fill = "gray95"), # facet label background
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank(),
        axis.text.y = element_text(color = "white")) +
  geom_segment(mapping = aes(x = start, y = 0, xend = end, yend = 0),
       size = 200, 
       colour = fillcolor,
       alpha = 0.2) +
  facet_wrap(vars(year), scales = "free_x", ncol = 1) +
  geom_blank(mapping = aes(x = x_min)) + # Adds the x_min column to force the x-axis to include the whole field season - it's just some invisible stuff in May and November to make the stacked plot look nice
  # also need to specify data down here now due to geom_vline 
  geom_blank(mapping = aes(x = x_max)) + 
  geom_vline(data = sampledatelist,
             inherit.aes = F, 
             mapping = aes(xintercept = sampledatelist), 
             color = "black") + # add each individual sample
  # scale break is breaking coord_cartesian(lim)
    # scale break is breaking coord_cartesian(lim)     https://github.com/YuLab-SMU/ggbreak/issues/59 put cartesian before
  scale_x_datetime(breaks = "month", date_labels = "%b-%y") + #, limits = as.POSIXct(c("2021-05-01", "2023-11-01"
  #coord_cartesian(ylim = c(-0.5, 0.5)) + # Zooms in the plot without deleting rectangle
  ylim(c(-0.01, 0.01)) +
  # clip = "off" allows data points to be drawn in the margins
  # expand = 0 doesn't add an expansion factor to the limits to keep the data off the axes
  ggtitle("PPS Deployment Windows") +
  xlab("Date") +
  ylab("") 

winplt_stk_mrk

ggsave(plot = winplt_stk_mrk,
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesLinesStacked.png"), 
           width = widthpx_stacked, 
           height = heightpx_stacked, 
           units = "px")
```


## Adding CTD

### CTD as lines

```{r}
# Graph 5: Shaded gray sampling windows + CTD lines
  

ggplot(data = stend, 
       mapping = aes(x = start, y = 0, xend = end, yend = 0)) +
  theme_bw() +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 45, hjust = 1), 
        strip.text = element_text(size = 12), 
        strip.background = element_rect(fill = "gray95"), # facet label background
        axis.text.x.top = element_blank(), # Needed to delete the extra axis created by ggbreak
        axis.ticks.x.top = element_blank(),
        axis.line.x.top = element_blank(),
        axis.text.y = element_text(color = "white")) +
  geom_segment(size = 200, 
       colour = fillcolor,
       alpha = 0.2) +
  geom_vline(data = ctddatelist, aes(xintercept = ctddatelist), color = ctdcolor) + # add CTD
  # scale break is breaking coord_cartesian(lim) https://github.com/YuLab-SMU/ggbreak/issues/59 put cartesian before
  scale_x_datetime(breaks = "month", date_labels = "%b-%y", limits = as.POSIXct(c("2021-05-01", "2023-11-01"))) +
  #coord_cartesian(ylim = c(-0.5, 0.5)) + # Zooms in the plot without deleting rectangle
  ylim(c(-0.01, 0.01)) +
  # clip = "off" allows data points to be drawn in the margins
  # expand = 0 doesn't add an expansion factor to the limits to keep the data off the axes
  ggtitle("PPS Deployment Windows") +
  xlab("Date") +
  ylab("") +
  scale_x_break(as.POSIXct(c("2021-11-01", "2022-05-01"))) +
  scale_x_break(as.POSIXct(c("2022-11-01", "2023-05-01"))) 

ggsave(filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesCTD.png"), 
           width = widthpx, 
           height = heightpx, 
           units = "px")
```


### CTD as dots

```{r}
# Graph 6: Shaded gray sampling windows with CTD as dots

CTD_dots <- geom_point(inherit.aes = F, data = ctddatelist, aes(x = ctddatelist, y = 0), color = ctdcolor) # make CTDs dots to distinguish, since they're ususally on the borders of deployment windows

winplot_ctd <- winplot+CTD_dots
winplot_ctd

ggsave(plot = winplot_ctd, 
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesCTDDots.png"), 
           width = widthpx, 
           height = heightpx, 
           units = "px")
```

### CTD + Stacked Years

```{r}
# Graph 7: Shaded gray sampling windows with CTD as dots + Stacked Years

# https://stackoverflow.com/questions/42588238/setting-individual-y-axis-limits-with-facet-wrap-not-with-scales-free-y
winplt_stk_ctd <- winplot_stk + CTD_add

winplt_stk_ctd

ggsave(plot = winplt_stk_ctd,
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesStackedCTD.png"), 
           width = widthpx_stacked, 
           height = heightpx_stacked, 
           units = "px")
```

### CTD + Each Sample Marked

```{r}
# Graph 8: Each sample marked + CTD
winplt_mrk_ctd <- winplt_mrk + CTD_add

winplt_mrk_ctd

ggsave(plot = winplt_mrk_ctd, 
       filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesLinesCTD.png"), 
           width = widthpx, 
           height = heightpx, 
           units = "px")
```


### CTD + Samps + Stacked Years

```{r}
# Graph 9: Each sample as a line + CTD + Stacked Years
winplt_stk_mrk_ctd <- winplt_stk_mrk + CTD_add

winplt_stk_mrk_ctd

ggsave(plot = winplt_stk_mrk_ctd, filename = here("Manuscript_Plots/Deployment_Window/DeploymentWindowsBoxesLinesStackedCTD.png"), 
           width = widthpx_stacked, 
           height = heightpx_stacked, 
           units = "px")
```

# Point plots

## Temp vs DO
```{r}
ggplot(envData, aes(x = temperature, y = DO)) +
  geom_point() +
    theme_bw() +
  theme(text = element_text(size = 20),
        axis.text.x = element_text(size = 25),
        axis.text.y = element_text(size = 25)
        ) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Oxygen vs Temperature", x = "Temperature (ºC)", y = "Dissolved Oxygen (Saturation %)")

ggsave(here("Manuscript_Plots", "Binomial_Regressions", "Oxygen_Temperature.png"), width = 2000, height = 2000, units = "px")
```

## Filter for species of interest

```{r, eval = T}
# later replace this with some vector from a spreadsheet
SOI <- c("Acartia longiremis", "Oncorhynchus kisutch", "Clupea pallasii", "Citharichthys sordidus")

eDNAxEnvData <- eDNAxEnvData %>% 
  filter(Species %in% SOI)
```

# Binomial Regression

**Binomial Regression Assumptions**

1. Independent observations - *Check!*
2. No multicollinearity: Continuous independent variables should not be correlated with each other - *NOT MET - shouldn't do a DO x Temp x Presence model*
3. Binomial dependent variable, with two **mutually exclusive** values - *Check!*
4. Continuous independent variable - *Check!*
5. Linearity between the binomial and continuous variables - this is checked with a logit transformation - *See below, this varies by species*
6. Sufficient obeservations - the source I found recommends > 15 - *Check! 10 should be alright but I will note it in the results*
7. No outliers - *One outlier, which appears to be part of a real heatwave - I have not filtered it out yet*


## Check for outliers

```{r}
ggplot(envData, aes(x = DO, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvData, mapping = aes(x = DO, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
ggsave(filename = here("Manuscript_Plots", "Binomial_Regressions", "SpeciesPresence_TempxDO.png"), width = 2500, height = 2000, units = "px")
```

## Filter out outliers & other cleaning 

```{r, eval = T}
eDNAxEnvData <- eDNAxEnvData %>% 
  filter(temperature < 11)

eDNAxEnvData <- eDNAxEnvData %>% 
  mutate(year = as.factor(year)) %>% # Clean for boxtidwell
  filter(!is.na(Present) & !is.na(DO) & !is.na(year)) %>%  # Clean for NA values
  filter(temperature < 11)
```

```{r}
ggplot(envData, aes(x = DO, y = temperature)) +
  geom_point(color = "gray90", alpha = 0.5) +
  geom_point(eDNAxEnvData, mapping = aes(x = DO, y = temperature, color = Present, shape = Present), inherit.aes = F) +
    scale_shape_manual(values = c(1, 19)) +
  theme_bw() +
  theme(text = element_text(size = 15), strip.text = element_text(size = 8), strip.background = element_rect(fill = "gray95")) +
  labs(title = "Dissolved Oxygen vs. Temp + Species Presence", y = "Temperature") +
  facet_wrap(facets = vars(Species))
```


## Check for linearity

"Assumption #5 involves the necessity of a linear relationship between the continuous independent variables and the logit transformation of the dependent variable. This linearity assumption implies that **for continuous independent variables** like income level, hours of exercise per week, and blood sugar levels, **there should be a linear relationship with the logit of the dependent variable**, such as the probability of developing diabetes. Various methods can be employed to assess this linearity, with one common approach being the **Box-Tidwell procedure.** This technique involves creating interaction terms between each continuous independent variable and its natural logarithm and adding these to the logistic regression model. This technique can be implemented using software like **SPSS Statistics, which offers the Binary Logistic procedure to test for this assumption.** The results of this test are then interpreted to decide the next steps in the analysis, depending on whether the linearity assumption holds or is violated. If the assumption is met, the analysis can proceed as planned. However, if the assumption is not met, adjustments to the model or alternative methods may be necessary to address the non-linearity appropriately." - [Binomial Logistic Regression](https://www.amstatisticalconsulting.com/binomial-logistic-regression/)


Basically, this equation needs to have a somewhat linear relationship with the independent variable: 

$$g(p) = log\left( \frac{p}{1-p} \right)$$
Where p = probability of 1 (if 1, p. if 0, 1-p)?


## Looped Binomial Regression

```{r}
# First, filter for more than 50 detections

Detect15Sum <- eDNAxEnvData %>% 
  group_by(Species) %>% 
  summarize(Detections = sum(Present)) %>% 
  arrange(desc(Detections)) %>% 
  filter(Detections > 15 & !is.na(Species)) 
Detect15Species <- Detect15Sum$Species # 37 species

Detect15 <- eDNAxEnvData %>% 
  filter(Species %in% Detect15Species) %>% 
  mutate(year = as.factor(year)) %>% # Clean for boxtidwell
  filter(!is.na(Present) & !is.na(DO) & !is.na(year))

Detect15Split <- split(Detect15, Detect15$Species) # Split by species

# Make a function that'll do the binomial regression AND put the results in a dataframe

SpeciesMod <- c()
InterceptP <- c()
SlopeP <- c()
SampleSizes <- c()

#modtest <- glm(Detect15Split[[1]], formula = Present ~ DO, family = "binomial")
#modsum <- summary(modtest)
#Coefficients: modsum$coeffifients[1 = intercept/2 = slope, 4]

for (i in 1:length(Detect15Split)) {
  SpName <- names(Detect15Split)[i]
  SpeciesMod <- c(SpeciesMod, SpName)
  
  df <- Detect15Split[[i]]
  
  SampleSize <- sum(df$Present)
  SampleSizes <- c(SampleSizes, SampleSize)
  
  # lapply(joinSpeciesList, glm, formula = Present ~ DO, family = "binomial")
  mod <- glm(df, formula = Present ~ DO, family = "binomial")
  modsum <- summary(mod)
  
  Intercept <- round(modsum$coefficients[1,4], digits = 5)
  Slope <- round(modsum$coefficients[2,4], digits = 5)
  InterceptP <- c(InterceptP, Intercept)
  SlopeP <- c(SlopeP, Slope)
}

BinomialRegressionResults <- data.frame(Species = SpeciesMod, InterceptP, SlopeP, n = SampleSizes)

SignificantSlopes <- BinomialRegressionResults %>% filter(SlopeP < 0.05)
SignificantIntercepts <- BinomialRegressionResults %>% filter(InterceptP < 0.05)

SigSp <- SignificantSlopes$Species
SigSpDf <- eDNAxEnvData %>% 
  filter(Species %in% SigSp)
```

## Box Tidwell Loop
```{r}
SigSpBox <- eDNAxEnvData %>%
  filter(!is.na(DO)) %>% 
  filter(Species %in% c("Aphanomyces astaci", "Aureococcus anophagefferens", "Thecamonas trahens", "Chaetoceros socialis", "Lithodesmium intricatum", "Cyphoderia ampulla", "Bathycoccus prasinos", "Clytia gregaria", "Paramoeba branchiphila", "Citharichthys sordidus",  "Thermocyclops inversus", "Pennaria disticha", "Clupea pallasii")) # Species I actually want to run this on, NA values removed

fail <- c("Acartia longiremis", "Chrysochromulina sp. NIES-1333", "Stephanopyxis turris", "Vannella sp.", "Heterosigma akashiwo", "Parvamoeba rugata") # Species where box-tidwell failed - they have DO NA values but the drop_na should fix that

LinearSplit <- split(SigSpBox, SigSpBox$Species) # Split by species, deleting species that broke the Box-Tidwell test for now

LinearSplit <- lapply(LinearSplit, drop_na, c(DO, Present, year))

test <- LinearSplit[[1]]
test2 <- LinearSplit[[2]]

# Do a box Tidwell test on everything that had significant slopes
# This only works on some of the species for some reason?
SpeciesBox <- c()
BoxTidwellP <- c()

for (i in 1:length(LinearSplit)) {
  print(paste(names(LinearSplit)[i], sep = " ", "Presence vs Oxygen"))
  box <- boxTidwell(Present ~ DO, 
                    other.x = ~ year, # it breaks if i don't do this idk
                    data = LinearSplit[[i]], 
                    na.action = na.exclude)
  
  x <- round(box$result[3], digits = 5)
  SpeciesBox <- c(SpeciesBox, names(LinearSplit)[i])
  BoxTidwellP <- c(BoxTidwellP, x)
}

BoxTidwellResults <- data.frame(SpeciesBox, BoxTidwellP) %>% arrange(desc(BoxTidwellP))

BoxTidwellSuccess <- BoxTidwellResults %>% filter(BoxTidwellP > 0.05)
```

# Graph binomial regression

# Loop to graph binom regresion
```{r}
AllSplit <- split(eDNAxEnvData, eDNAxEnvData$Species)
# Add title
# Add better axis labels
for (i in 1:length(AllSplit)) {
  df <- AllSplit[[i]] %>% mutate(PresentBin = case_when(Present == T ~ 1, Present == F ~ 0)) %>%  # Select one species and mutate to 0/1 so that the regression will run
    filter(temperature < 11) # Filter out the outlier
  
  species <- as.character(AllSplit[[i]][1,7]) # Extract species name for title + export filename

  ggplot(df, aes(x = DO, y = PresentBin, color = PresentBin)) +
    geom_point(size = 4) +
    scale_color_gradient(high = "dodgerblue3", low = "orange2", guide = "legend", na.value = "black") + # Color to match presence/absence graphs
    geom_smooth(method = "glm", method.args = list(family = "binomial"), se = F) +
    # Print the binomial regression
    stat_poly_eq(use_label(c("R2", "P")), size = 6, label.y = 0.75, label.x = 60) + # Print R^2 and p-value from the model
    theme_bw() +
    theme(text = element_text(size = 20)) +
    labs(title = species, x = "Oxygen Saturation (%)", y = "Species Present?")
  
    ggsave(filename = here("Manuscript_Plots", "Binomial_Regressions",
                           paste(species, sep = ".", "png")), # Save graph
           width = 2300, 
           height = 2000, 
           units = "px")
}

system("say Binomial Regression Graphs Complete!")
```

To kill the talking enter in terminal `while true; do pkill -9 say; done`